Normalization by Evaluation for Call-By-Push-Value and Polarized Lambda Calculus

### 2019-05-09 ppdp19/ppdp19.tex
[ ppdp19 ] proof-read rest of 2.3 and 2.4
- added informal explanation of the cover monad, linking to oracle
- added implicit context bindings as subscripts
- runNf can be restricted to positive types P
- added conclusion to 2.3 and motivation of 2.4

\begin{remark}
While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^P           & u & = & \tabort\;u \\
% \end{array}
% \qquad
% \begin{array}{l@{~}lcl}
%   \\
  \tabort^{A_1 \times A_2} & u & = & \tpair\,(\tabort^{A_1}\,u)\,(\tabort^{A_2}\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
\end{array}
% \begin{array}{l@{~}lcl}
%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
%   \tabort^{\tyone}     & u & = & \tunit \\
%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
%   \tabort^P           & u & = & \tabort\;u \\
% \end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.
\end{remark}

@@

Note also that in the equation for $\up{A \To B}$,
the neutral $u : \Ne {(A \To B)} \Gamma$ is
transported into $\Ne {(A \To B)} \Delta$ via reindexing with
$\tau : \Gamma \sublist \Delta$, in order to be applicable to the normal form
$\down[\Delta]A a$
reified from the semantic value $a$.

A direct extension of our presheaf semantics to positive types cannot
work.  For instance, with $\den \tyzero = \hat0$, simply
% $\up[\tyzero] \tyzero \vv0 : \bzero$
$\tfresh^{\tyzero}_{\cempty} : \bzero$
would give us an
inhabitant of the empty set, which means that reflection at the empty
type would not be definable.  Similarly, the setting
$\den{A + B} = \den A \hatplus \den B$ is refuted by
% $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
$\tfresh^{A + B}_{\cempty} : \den A (A{+}B) \bplus \den B (A{+}B)$
% $\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
which would require us to make a decision of whether $A$ holds or $B$ holds
while only be given a hypothesis of type $A + B$.
Not even the usual interpretation of base types $\den o = \tNe\,o$
works in the presence of sums, as we would not be able to interpret
the term
$\tabs\,(\tcase\,\vv0\,\vv0\,\vv0) :
(o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
What is needed are case distinctions on neutrals in the semantics,
allowing us the elimination of positive hypotheses before producing a
semantic value,
and we shall capture this capability in a strong monad $\C$
which can \emph{cover} the cases.

@@

We are looking for a cover monad $\C$ that
offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{l@{~}c@{~}l@{\quad}l}
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
%\\ & & \mbox{case on absurd neutral}
\\
\tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
\\
  && {} \to \C\,\B\,(\cext \Gamma {A_1})
  \to \C\,\B\,(\cext \Gamma {A_2})
  \to \C\,\B\,\Gamma
%\\ & & \mbox{case on neutral}
\\
\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
%\\ & & \mbox{run the monad ($\tNf$ only)}
\\
\end{array}
\]
To make things concrete,
we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined
%inductively by
as an inductive family with constructors
$\treturn[\Cov]$, $\tabort^{\Cov}$, and
$\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
and $A_2$, resp.  Leaves are either
labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind
(aka Kleisli extension) replaces
these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$
which flattens a 2-level case tree, \ie, a case tree with case trees as
leaves, into a single one.
Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
and $\tabort^{\Cov}$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

The unwinding of the monad $\C$ via $\trunNf$ can be extended to a
natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
type $A$, meaning we can \emph{run} the monad,
pushing its effects into $\den A$.%
\footnote{In categorical terminology, the existence of
$\trun^A$ means that
all semantic types $\den{A}$ fulfill the \emph{weak sheaf condition},
aka \emph{weak pasting}.}
% All semantic types fulfill the weak sheaf condition aka weak pasting,
% meaning there is
% a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
% type $A$.  In other words, we can \emph{run} the monad, pushing its
% effects into $\den A$.
We proceed by induction on $A$.
Positive semantic types $\den P$ are already monadic,
and $\trun^P$ is simply the $\tjoin$ of the monad $\C$.
At negative types we can recurse pointwise at a smaller type,
exploiting that values of negative types are essentially
(finite or infinite) tuples.
\[
\begin{array}{ll@{~~}cl}%{l@{~}l@{~~}c@{~~}l}
  \multicolumn 4 l {\trun^A ~:~ \C \den A \todot \den A} \\
  \trun^{\tyone} & c & = & \ttempty
    \\
  \trun^{A \times B} & c & = &
      (\trun^A\,(\tmap[\C]\;\pi_1\;c),\
       \trun^B\,(\tmap[\C]\;\pi_2\;c))
  \\
  \trun^{A \To B} & c  & = & \lambda \,\tau \; a. \ \trun^B\,
      (\stmap[\C]\;(\lambda\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))
  \\ &&& \qquad \qquad (\tren\;\tau\;c))
  \\
  \trun^{o}      & & = & \tjoin[\C]
  \\
  \trun^{\tyzero} & & = & \tjoin[\C]
  \\
  \trun^{A + B}   & & = & \tjoin[\C]
  \\
\end{array}
\]
For the case of function types $A \To B$, we require the monad $\C$ to be
\emph{strong}, which amounts to having
$\stmap[\C]_\Gamma\, \ell : \C\A\,\Gamma \to \C\,\B\,\Gamma$ already for
a ``local'' presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
The typings are
$c : \C \den{A \To B} \Gamma$ and
$\tau : \Gamma \sublist \Delta$ and
$a : \den A \Delta$, and now we want to apply every function
$f : \den{A \To B}$ in the cover $c$ to argument $a$.  Clearly,
$\tmap[\C]$ is not applicable since it would expect a
global presheaf morphism
$\den{A \to B} \todot \den B$, \ie, something that works in \emph{any}
context.  However, applying to $a : \den A \Delta$ can only work in
context $\Delta$ or any extension $\tau' : \Delta \sublist \Phi$,
since we can transport $a$ to such a $\Phi$ via
$a' := \tren\; \tau'\; a : \den A \Phi$ but not to a context unrelated to
$\Delta$.  We obtain our input to $\trun^B$ of type $\C \den B \Gamma$
as an instance of $\stmap[\C]_\Gamma$ applied to the local presheaf morphism
$(\lambda \; \tau'\, f. \  f\;\tid\;a') :
\Delta \sublist \Phi \to  \den{A \To B}\Phi \to \den B \Phi$
% (\tau' : \Delta \sublist \Phi) (f : \den{A \To B}\Phi) \to \den B \Phi$
and the transported cover $\tren\;\tau\;c : \C\den{A \To B}\Delta$.

We extend the type interpretation pointwise to contexts, \ie,
$\den \cempty = \hatone$ and
$\den{\cext \Gamma A} = \den \Gamma \hattimes \den A$
and obtain a natural projection function
$\tlookup\,(x : A \in \Gamma) : \den \Gamma \todot \den A$
from the semantic environments.
% \[
% \begin{array}{lcl}
%   \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\
%   \tlookup\;\tzero     & = & \pi_2 \\
%   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
% \end{array}
% % \begin{array}{lcl}
% %   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\
% %   \tlookup\;\tzero     & = & \pi_2 \\
% %   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
% % \end{array}
% \]
The evaluation function
\fbox{$\denty {t : A \from \Gamma}  : \den \Gamma \todot \den A$}
% \fbox{$\dent {t : A \from \Gamma} \_ : \den \Gamma \todot \den A$}
can now be defined by recursion on $t$.
% using an auxiliary definition
% \fbox{$\fden{t : B \from \cext \Gamma A} : \den \Gamma \todot \den{A \To B}$}.
Herein, the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
% Herein, $\gamma : \den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
\[
\begin{array}{lcl}
  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
  \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
  \dent{\tapp\,t\,u} \gamma & = &
     \dent t \gamma\;\tid\;\dent u \gamma \\
  \dent{\tunit}\gamma & = & \ttempty \\
  \dent{\tpair\,t_1\,t_2}\gamma & = &
    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
  \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
  \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
     \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
  \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
\end{array}
\]
For the interpretation of the binders $\tabs$ and $\tcase$ we use the
mutually defined $\fden{\_}$.
\[
\begin{array}{l}
  \fden{t : B \from \cext \Gamma A}
    : \den \Gamma \todot \den{A \To B} \\
    \quad =
      \lambda\, (\gamma : \den \Gamma \Delta)
      \,(\tau : \Delta \sublist \Phi)
      \,(a : \den A \Phi)
      .\ \,
      \dent t {(\tren\,\tau\,\gamma,\ a)} \\
  % \fden{\_} ~~:~~ B \from \cext \Gamma A} \to \den \Gamma \todot \den{A \To B}
  % \\
  % \fdent{t : B \from \cext \Gamma A}\gamma
  %     \,(\tau : \Delta \sublist \Phi)
  %     \,(a : \den A \Phi)
  %     ~~=~~
  %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\
\end{array}
\]
The coproduct eliminations $\denabort$ and $\dencase$
targeting an arbitrary semantic type $\den B$ are definable thanks to
the weak sheaf property, \ie, the presence of pasting via $\trun^B$
for any type $B$, and strong functoriality of $\C$.
\[
\begin{array}{l@{~}c@{~}l}
  \denabort^B & : & \den \tyzero \todot \den B \\
  \denabort^B & = & \trun^B \comp \tmap[\C]\,\tmagic
\\[2ex]
  \dencase^B  & : & \den{A_1 + A_2}
    \todot \den{A_1 \To B}
    \todot \den{A_2 \To B}
    \todot \den{B} \\
  \dencase^B & = & \lambda\,c\;f_1\;f_2.\
    \trun^B (\stmap[\C]\;(\lambda\,\tau.\ [f_1\,\tau,\ f_2\,\tau])\;c) \\
\end{array}
\]
To complete the normalization function
$\tnorm\,(t : A \from \Gamma) = \downG A \dent t {\idenv \Gamma}$
we define the identity environment $\idenv \Gamma : \den \Gamma \Gamma$,
which maps each free index to its corresponding unknown in the
semantics, by recursion on $\Gamma$:
\[
\begin{array}{lcl}
  \idenv \cempty & = & \ttempty \\
  \idenv {\cext \Gamma A} & = & (\tren\;\twk^A\,\idenv\Gamma, \  \tfresh^A_\Gamma)\\
\end{array}
\]

@@

Besides the free cover monad $\Cov$, there is another canonical
implementation of its interface, the continuation monad.
Filinski
\cite[Section~5.4]{filinski:semaccounttdpe}
\cite[Section~3.2]{filinski:tlca01}
already utilized the continuation monad for
normalization by evaluation.
In our setting, we use a continuation monad $\CC$ on presheaves
defined as
% An alternative implementation of $\C$ is the continuation monad $\CC$
% on presheaves as, for instance, employed by Ilik et al.\
% \cite{ilikLeeHerbelin:apal10}
% for Kripke models and NbE of classical logic.
\[
  \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
  .
\]
The answer type of this continuation monad is always $\tNf$, however,
we are polymorphic in the simple type $A$ of normal forms we produce.%

>>>>

\begin{remark}[Positive eliminations into negative types]
\label{rem:abort}
While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^P           & u & = & \tabort\;u \\
% \end{array}
% \qquad
% \begin{array}{l@{~}lcl}
%   \\
  \tabort^{A_1 \times A_2} & u & = & \tpair\,(\tabort^{A_1}\,u)\,(\tabort^{A_2}\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
\end{array}
% \begin{array}{l@{~}lcl}
%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
%   \tabort^{\tyone}     & u & = & \tunit \\
%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
%   \tabort^P           & u & = & \tabort\;u \\
% \end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.
\end{remark}

@@

Note also that in the equation for $\up{A \To B}$,
the neutral $u : \Ne {(A \To B)} \Gamma$ is
transported into $\Ne {(A \To B)} \Delta$ via reindexing with
$\tau : \Gamma \sublist \Delta$, in order to be applicable to the normal form
$\down[\Delta]A a$
reified from the semantic value $a : \den A \Delta$.

A direct extension of our presheaf semantics to positive types cannot
work.  For instance, with $\den \tyzero = \hat0$, simply
% $\up[\tyzero] \tyzero \vv0 : \bzero$
$\tfresh^{\tyzero}_{\cempty} : \bzero$
would give us an
inhabitant of the empty set, which means that reflection at the empty
type would not be definable.  Similarly, the setting
$\den{A + B} = \den A \hatplus \den B$ is refuted by
% $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
$\tfresh^{A + B}_{\cempty} : \den A (A{+}B) \bplus \den B (A{+}B)$
% $\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
which would require us to make a decision of whether $A$ holds or $B$ holds
while only be given a hypothesis of type $A + B$.
Not even the usual interpretation of base types $\den o = \tNe\,o$
works in the presence of sums, as we would not be able to interpret
the term
$\tabs\,(\tcase\,\vv0\,\vv0\,\vv0) :
(o+o) \To o$ in our semantics, because $\Ne o (o+o)$ is empty.
What is needed are case distinctions on neutrals in the semantics,
allowing us the elimination of positive hypotheses before producing a
semantic value,
and we shall capture this capability in a strong monad $\C$
which can \emph{cover} the cases.

@@

We are looking for a monad $\C$, called a \emph{cover monad}, that
takes the role of the \emph{oracle} alluded to in the introduction,
and offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{l@{~}c@{~}l@{\quad}l}
\trunNf[\C] & : & \C\,(\tNf\, P) \todot \tNf\, P
%\\ & & \mbox{run the monad ($\tNf$ only)}
\\
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
%\\ & & \mbox{case on absurd neutral}
\\
\tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
\\
  && {} \to \C\,\B\,(\cext \Gamma {A_1})
  \to \C\,\B\,(\cext \Gamma {A_2})
  \to \C\,\B\,\Gamma
%\\ & & \mbox{case on neutral}
\\
\end{array}
\]
Method $\trunNf[\C]$ enables us to \emph{run} a monadic computation of
a normal form, $\C\,(\tNf\,P)\,\Gamma$, and delivers a normal form
$\Nf P \Gamma$.  However, $\trunNf[\C]$ is not just the inverse of
$\treturn[\C]$; instead, the information contained in $\C$ is reified and
becomes part of the normal form delivered by $\trunNf[\C]$.

Method $\tabort^\C_\Gamma$ allows us to end a computation by exhibiting an
inconsistency witnessed by a neutral term $u : \Nf A \Gamma$.
Remember that, ultimately, computations in $\C$ produce a normal form
extracted by $\trunNf[\C]$.  With $\tabort^\C$ we notify the oracle
that we are in an absurd case and the desired normal form can be
trivially constructed by the $\tabort$-function of \Cref{rem:abort}.

Method $\tcase^\C_\Gamma$ allows us, in a computation
$\C\,\B\,\Gamma$,
to ask the oracle about a neutral
term $\NeP{A_1 + A_2}\Gamma$ of sum type.  We have to supply
handlers for both possible answers:
A computation $\C\,\B\,(\cext\Gamma{A_1})$ that can utilize an
additional hypothesis of type $A_1$ in the case of a left injection,
and analogously a computation $\C\,\B\,(\cext\Gamma{A_2})$ for the
case of a right injection.

The similarity of the monad services $\tabort^\C$ and $\tcase^\C$ to
the corresponding constructors for normal forms, or their
generalization of \Cref{rem:abort}, is hard to miss.
Unsurprisingly, just \emph{case trees} are a first instance of a cover monad,
% To make things concrete,
% we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined
%inductively by
as an inductive family with constructors
$\treturn[\Cov]$, $\tabort^{\Cov}$, and
$\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by neutral terms
of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
and $A_2$, resp.  Leaves are either
labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind
(aka Kleisli extension) replaces
these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$
which flattens a 2-level case tree, \ie, a case tree with case trees as
leaves, into a single one.
Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
and $\tabort^{\Cov}$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

The unwinding of the monad $\C$ via $\trunNf$ can be extended to a
natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
type $A$, meaning we can \emph{run} the monad,
pushing its effects into $\den A$.%
\footnote{In categorical terminology, the existence of
$\trun^A$ means that
all semantic types $\den{A}$ fulfill the \emph{weak sheaf condition},
aka \emph{weak pasting}.}
% All semantic types fulfill the weak sheaf condition aka weak pasting,
% meaning there is
% a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
% type $A$.  In other words, we can \emph{run} the monad, pushing its
% effects into $\den A$.
We proceed by induction on $A$.
Positive semantic types $\den P$ are already monadic,
and $\trun^P$ is simply the $\tjoin$ of the monad $\C$.
At negative types we can recurse pointwise at a smaller type,
exploiting that values of negative types are essentially
(finite or infinite) tuples.
\[
\begin{array}{l@{~~}l@{~~}c@{~~}l}%{l@{~}l@{~~}c@{~~}l}
  \multicolumn 4 l {\trun^A ~:~ \C \den A \todot \den A} \\
  \trun^P & c & = & \tjoin[\C]\;c \\
  \trun^{\tyone} & c & = & \ttempty
    \\
  \trun^{A \times B} & c & = &
      (\trun^A\,(\tmap[\C]\;\pi_1\;c),\
       \trun^B\,(\tmap[\C]\;\pi_2\;c))
  \\
  \trun^{A \To B} & c  & = & \lambda\,_\Delta \,\tau \; a. \ \trun^B\,
      (\stmap[\C]\;(\lambda\,_\Phi\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))
  \\ &&& \qquad \qquad (\tren\;\tau\;c))
  \\
  % \trun^{o}      & & = & \tjoin[\C]
  % \\
  % \trun^{\tyzero} & & = & \tjoin[\C]
  % \\
  % \trun^{A + B}   & & = & \tjoin[\C]
  % \\
\end{array}
\]
For the case of function types $A \To B$, we require the monad $\C$ to be
\emph{strong}, which amounts to having
$\stmap[\C]_\Gamma\, \ell : \C\A\,\Gamma \to \C\,\B\,\Gamma$ already for
a ``local'' presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
The typings are
$c : \C \den{A \To B} \Gamma$ and
$\tau : \Gamma \sublist \Delta$ and
$a : \den A \Delta$, and now we want to apply every function
$f : \den{A \To B}$ in the cover $c$ to argument $a$.  Clearly,
$\tmap[\C]$ is not applicable since it would expect a
global presheaf morphism
$\den{A \to B} \todot \den B$, \ie, something that works in \emph{any}
context.  However, applying to $a : \den A \Delta$ can only work in
context $\Delta$ or any extension $\tau' : \Delta \sublist \Phi$,
since we can transport $a$ to such a $\Phi$ via
$a' := \tren\; \tau'\; a : \den A \Phi$, but not to a context unrelated to
$\Delta$.  We obtain our input to $\trun^B$ of type $\C \den B \Gamma$
as an instance of $\stmap[\C]_\Gamma$ applied to the local presheaf morphism
$(\lambda\,_\Phi \; \tau'\, f. \  f\;\tid\;a') :
\Delta \sublist \Phi \to  \den{A \To B}\Phi \to \den B \Phi$
% (\tau' : \Delta \sublist \Phi) (f : \den{A \To B}\Phi) \to \den B \Phi$
and the transported cover $\tren\;\tau\;c : \C\den{A \To B}\Delta$.

We extend the type interpretation pointwise to contexts, \ie,
$\den \cempty = \hatone$ and
$\den{\cext \Gamma A} = \den \Gamma \hattimes \den A$
and obtain a natural projection function
$\tlookup\,(x : A \in \Gamma) : \den \Gamma \todot \den A$
from the semantic environments.
% \[
% \begin{array}{lcl}
%   \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\
%   \tlookup\;\tzero     & = & \pi_2 \\
%   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
% \end{array}
% % \begin{array}{lcl}
% %   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\
% %   \tlookup\;\tzero     & = & \pi_2 \\
% %   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
% % \end{array}
% \]
The evaluation function
\fbox{$\denty {t : A \from \Gamma}  : \den \Gamma \todot \den A$}
% \fbox{$\dent {t : A \from \Gamma} \_ : \den \Gamma \todot \den A$}
can now be defined by recursion on $t$.
% using an auxiliary definition
% \fbox{$\fden{t : B \from \cext \Gamma A} : \den \Gamma \todot \den{A \To B}$}.
Herein, the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
% Herein, $\gamma : \den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
\[
\begin{array}{lcl}
  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
  \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
  \dent{\tapp\,t\,u} \gamma & = &
     \dent t \gamma\;\tid\;\dent u \gamma \\
  \dent{\tunit}\gamma & = & \ttempty \\
  \dent{\tpair\,t_1\,t_2}\gamma & = &
    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
  \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
  \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
     \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
  \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
\end{array}
\]
For the interpretation of the binders $\tabs$ and $\tcase$ we use the
mutually defined $\fden{\_}$.
\[
\begin{array}{l}
  \fden{t : B \from \cext \Gamma A}
    : \den \Gamma \todot \den{A \To B} \\
    \quad =
      \lambda\,_\Delta\, (\gamma : \den \Gamma \Delta)
      \,_\Phi
      \,(\tau : \Delta \sublist \Phi)
      \,(a : \den A \Phi)
      .\ \,
      \dent t {(\tren\,\tau\,\gamma,\ a)} \\
  % \fden{\_} ~~:~~ B \from \cext \Gamma A} \to \den \Gamma \todot \den{A \To B}
  % \fdent{t : B \from \cext \Gamma A}\gamma
  %     \,(\tau : \Delta \sublist \Phi)
  %     \,(a : \den A \Phi)
  %     ~~=~~
  %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\
\end{array}
\]
The coproduct eliminations $\denabort$ and $\dencase$
targeting an arbitrary semantic type $\den B$ are definable thanks to
the weak sheaf property, \ie, the presence of pasting via $\trun^B$
for any type $B$, and strong functoriality of $\C$.
\[
\begin{array}{l@{~}c@{~}l}
  \denabort^B & : & \den \tyzero \todot \den B \\
  \denabort^B & = & \trun^B \comp \tmap[\C]\,\tmagic
  \dencase^B  & : & \den{A_1 + A_2}
    \todot \den{A_1 \To B}
    \todot \den{A_2 \To B}
    \todot \den{B} \\
  \dencase^B & = & \lambda\,c\,f_1\,f_2.\
    \trun^B (\stmap[\C]\;(\lambda\,\tau.\ [f_1\,\tau,\ f_2\,\tau])\;c) \\
\end{array}
\]
To complete the normalization function
$\tnorm\,(t : A \from \Gamma) = \downG A \dent t {\idenv \Gamma}$
we define the identity environment $\idenv \Gamma : \den \Gamma \Gamma$,
which maps each free index to its corresponding unknown in the
semantics, by recursion on $\Gamma$:
\[
\begin{array}{lcl}
  \idenv \cempty & = & \ttempty \\
  \idenv {\cext \Gamma A} & = & (\tren\;\twk^A\,\idenv\Gamma, \  \tfresh^A_\Gamma)\\
\end{array}
\]
We have obtained a computable function $\tnorm$ that yields for any
derivation $A \from \Gamma$ a normal derivation $\Nf A \Gamma$.
This amounts to proving
consistency and canonicity of intuitionistic propositional logic, the
Curry-Howard counterpart of our STLC with products and sums.
The normalization function is parametrized by an arbitrary cover monad
$\C$, which can be implemented via case trees ($\Cov$).  However,
different implementations are possible.  In the next section, we exhibit
another canonical choice for $\C$: the parameterized continuation monad on presheafs.

@@

Besides the free cover monad $\Cov$, there is another canonical
implementation of its interface, the continuation monad.
Filinski
\cite[Section~5.4]{filinski:semaccounttdpe}
\cite[Section~3.2]{filinski:tlca01}
already utilized the continuation monad for
normalization by evaluation.
In our setting, we use a continuation monad $\CC$ on \emph{presheaves}
$\B$ defined by
% An alternative implementation of $\C$ is the continuation monad $\CC$
% on presheaves as, for instance, employed by Ilik et al.\
% \cite{ilikLeeHerbelin:apal10}
% for Kripke models and NbE of classical logic.
\[
  \CC\;\B = \forall P.\ (\B \hatto \tNf\,P) \hatto \tNf\,P
  .
\]
The answer type of this continuation monad is always $\tNf$, however,
we are polymorphic in the positive type $P$ of normal forms we produce.%

Agda has been really helpful to produce the rather technical but
straightforward evidence that $\CC$ is a strong monad.
The method $\trunNf[\CC] : \CC\,(\tNf\,P) \todot \tNf\,P$
exists by definition, using the identity continuation $\tNf\,P \hatto \tNf\,P$.
In the following, we demonstrate that $\CC$
enables matching on neutrals:
\[
\renewcommand{\left}{}
\renewcommand{\right}{}
\begin{array}{l}
  \tabort^{\CC}_\Gamma \, (u : \Ne \tyzero \Gamma) ~:~  \CC\,\B\,\Gamma \\[0.2ex]
  \tabort^{\CC}_\Gamma
    \; u
    \, _\Delta
    \, (\tau : \Gamma \sublist \Delta)
    \; (k : (\B \hatto \tNf\,P)\,\Delta)
     ~=~ \\  \qquad
    \tabort\,(\tren\,\tau\,u)
\\[2ex]
  \tcase^{\CC}_\Gamma
     \; (u : \NeP {A_1 + A_2} \Gamma) \, _\Delta \\ \qquad \quad
     \; (c_1 : \CC\,\B\,(\cext \Gamma {A_1}))
     \; (c_2 : \CC\,\B\,(\cext \Gamma {A_2}))
    ~:~ \CC\,\B\,\Gamma \\
  \tcase^{\CC}_\Gamma
    \; u
    \; c_1
    \; c_2
    \; (\tau : \Gamma \sublist \Delta)
    \; (k : (\B \hatto \tNf\,P)\,\Delta)
     ~=~ \\  \qquad
    \tcase\;(\tren\,\tau\,u)\;n_1\;n_2
    \\ \qquad \bwhere \\
\qquad
  \begin{array}{l@{~~}c@{~~}l}
     n_i & : & \Nf P {(\cext \Delta {A_i})} \\
     n_i & = & c_i
      \; (\tlift^{A_i}\,\tau
            : \cext \Gamma {A_i} \sublist \cext \Delta {A_i})
\\ &&
      \; \left(\lambda
            \, _\Phi
            \, (\tau' : \cext \Delta {A_i} \sublist \Phi)
            \; (j :  \B\,\Phi) .
            \ k
            \; (\twk^{A_i} \rcomp \tau')
            \; j
         \right)
  \end{array}
\end{array}
\]
The NbE algorithm using $\CC$ is comparable to Danvy's type-directed
partial evaluation \cite[Figure~8]{danvy:popl96}.  However, he uses
shift-reset style continuations, which can be programmed in the
continuation monad, and relies on Scheme's \texttt{gensym} to produce
fresh variables names rather than using the Kripke function space / presheaves.
